F-stats vs PCA

### 2021-04-16 fstats_pca.tex

	Notice that the entries of the covariance matrix $\MK$ are
	\begin{equation}
	k_{ij} = \sum_l y_{il} y_{jl} = \sum_l (x_{il} - \mu_l)(x_{yl} - \mu_l) = F_3(\MEAN; X_i, X_j)\text{,}
	\end{equation}
	where $\MEAN = (\mu_1, \dots \mu_k)$ is the vector of the mean allele frequency at each SNP. One interpretation is that $\MEAN$ denotes a ``pivot''-population whose allele frequency vector is the sample mean of the allele frequency space. Because of this, it will always lie at the origin of the PC-space, and performing a PCA is equal to a rotation around this ``pivot''-population.

	One implication of this is that $\MEAN$ strongly depends on sample composition. If we add many closely related populations (or, in an individual based PCA, many individuals from the same population), the PCA will be skewed disproportionally towards this cluster.

>>>>

	Notice that the entries of the covariance matrix $\MK$ are
	\begin{equation}
	k_{ij} = \sum_l y_{il} y_{jl} = \sum_l (x_{il} - \mu_l)(x_{yl} - \mu_l) = F_3(\MEAN; X_i, X_j)\text{,}
	\end{equation}
	where $\MEAN = (\mu_1, \dots \mu_k)$ is the vector of the mean allele frequency at each SNP. One interpretation is that $\MEAN$ denotes a ``pivot''-population whose allele frequency vector is the sample mean of the allele frequency space. By definition, it lies at the origin of the PC-space, and performing a PCA is equal to a rotation around this ``pivot''-population.

	One implication of this is that $\MEAN$ strongly depends on sample composition. If we add many closely related populations (or, in an individual based framework, )This yields an (informal) analogy to the midpoint rooting of a tree in phylogenetics \citep[see e.g.][]{felsenstein2004}. In the absence of any further information, a sensible choice for the root of a tree is the point that is furthest away from all the tips.

	Thus, we can also think of a PCA as performing a decomposition of an Outgroup-$F_3$-matrix, where we set the output to the mean of all marker allele frequencies. This is good practice if we do not have any outgroups available. However, we often also have outgroup data; for example when studying early European variation, Africans are sometimes a suitable outgroup. In this case, it would be consequential to perform an eigendecomposition of a true outgroup $F_3$-matrix.
